---
title: "Week2 Exercise"
author: "Matt Ming"
date: "1/27/2022"
output: html_document
---

This week's programming exercise:

Create a new repo on GitHub (can call it "Week2-Exercise" or something similar)

Copy this repo into Rstudio

Create an .Rmd file in Rstudio containing notes and exercises from today's class

Knit the .Rmd file

Commit and push the .Rmd, .html, and .Rproj files to the newly created repo

Share the repo w/ Dr. Di Fiore as a url via canvas and as a contributor on GitHub
(username is difiore)

# Class notes

## Data structures and getting data into R

### Using libraries

To get some example data, we first downloaded both a .xlsx and .csv file from
Dr. Di Fiore's ada-2022-datasets repo

We then loaded in several packages

```{r}
library(tidyverse)
library(readxl)
library(data.table)
```

Note that it is possible that your packages are out of date, in which case you'll
want to update them using the "packages" tab in Rstudio or just using install.packages()

### Matrices

Matrices are one way to store data

They must contain all one type of data (e.g., all numbers or all characters)

They fill either by col (default) or by row (set byrow=TRUE)

```{r}
matrix(1:30,nrow = 5,ncol = 6)
matrix(1:30,nrow = 5,ncol = 6,byrow = TRUE)
```

Indices in R can be called using [], and different dimensions (such as rows and
columns in a matrix) can be separated by a comma [n,m]

The result of the index call can be assigned to a variable

Indices can be set as ranges [a:b,c:d]

You can take all of the rows in some column by leaving the row part of the index
blank [,m] and same for getting all columns of some rows [n,]

```{r}
m <- matrix(1:30,nrow = 5,ncol = 6)
i <- m[1,1]
i
m[3:4,4:6]
m[,5]
m[1:3,]
```

### Lists and Dataframes

Lists and Dataframes can also store data but these are able to hold multiple types
of data

In Lists, the rows do not all need to have the same length

```{r}
s <- c("this", "is", "a", "vector", "of", "strings")
# this is a vector of character strings
m <- matrix(data=1:40, nrow=5, ncol=8) # this is a matrix
b <- FALSE # this is a boolean variable
l <- list(s, m, b)
l <- list(string = s,matrix = m,bool = b)
l
```

Dataframes are perhaps the most commonly used type of data structure as they are
basically lists but every vector within the list must have the same number of
elements, essentially creating tabular data

```{r}
df <-
  data.frame(
    firstName = c("Rick", "Negan", "Dwight", "Maggie", "Michonne"),
    community = c("Alexandria", "Saviors", "Saviors", "Hiltop", "Alexandria"),
    sex = c("M", "M", "M", "F", "F"),
    age = c(42, 40, 33, 28, 31)
  )
df
```

It's crucial to understand dataframes because reading in tables to R will auto-
matically convert the data to a dataframe

To import data from an external file, can use "read.csv" or "read.delim"

Note that when calling the file, you may need to specify the exact path to that
file, depending on where the file is saved

```{r}
df <- read.csv(file = "CPDS-1960-2014-reduced.csv",
               sep = ",",
               header = TRUE,
               stringsAsFactors = FALSE)
head(df)
```

Note that the packages tidyverse and readxl have their own functions for reading
in data from files

In tidyverse, use read_csv()

In readxl, use read_excel()

Both have their own variables so make sure to read the help notes
